Density-dependent selection and the limits of relative fitness

### 2018-03-15 manuscript.tex
l

For haploids growing in discrete time steps, the change in the abundance $n_i$ of type $i$ over a time step can be expressed as $\Delta n_i=(W_i - 1)n_i$ where $W_i$ is absolute fitness. The corresponding change in frequency is $\Delta p_i=\left(\frac{W_i}{\overline{W}}-1\right) p_i$, where $\overline{W}=\sum_i W_i p_i$. In continuous time, the Malthusian parameter $r_i$ replaces $W_i$ and we have $\frac{d n_i}{dt}=r_in_i$ and $\frac{d p_i}{dt}=(r_i-\overline{r}) p_i$ \citep[pp. 26]{crow_1970}. In the particular case that two types are present, a mutant labeled $1$ and a wildtype labeled $2$ for instance, then the continuous time equation for the mutant takes the familiar form
\begin{equation}
\frac{d p_1}{dt}=s p_1(1-p_1), \label{eq:canonical}
\end{equation}
where $s=r_1-r_2$ is the selection coefficient. Note that in the discrete case we can replace the $W_i$ with any set of values proportional to the $W_i$ and the changes in frequency $\Delta p_i$ will be unaffected. These values, which give no information about the dynamics of total population density $N=\sum_i n_i$, are called ``relative fitnesses'' \citep[pp. 468]{barton_2007}. Similarly, the selection coefficient and frequency dynamics in Eq.~\eqref{eq:canonical} would be unaffected if we were to add an arbitrary constant to both $r_1$ and $r_2$ (these would be relative log fitnesses).

In the absence of crowding, each type grows exponentially at a constant ``instrinsic'' rate $r_i$ (for concreteness, we will primarily use Eq.~\eqref{eq:canonical} as our point of comparison but similar logic applies in the discrete time case). The selection coefficient $s$ is then a constant equal to the difference in intrinsic population growth rates. The situation is more complicated when we allow for population crowding. Since crowded and uncrowded conditions can favor quite different traits \citep{travis_2013}, the Malthusian parameters $r_i$ may respond differently to changes in population density $N$, in which case the selection coefficient will be density-dependent. Eq.~\eqref{eq:canonical} would then need to supplemented with a model for the dynamics of $N$ to describe how the allele frequencies will change over time. The density and type frequency dynamics will in general be coupled \citep{travis_2013}. Thus, in a general crowded population the simple procedure of assigning constant relative fitness values to different types has to be replaced with an ecological description of absolute growth rates --- a substantially more complicated problem. Note that frequency-dependent selection does not raise such a severe problem, because we still only need to model the type frequencies, and not also the ecological variable $N$.

In practice, many population genetics models assume that $N$ has reached some fixed equilibrium value, and each type is assigned a constant relative fitness. Selection is typically assumed to operate through viability (e.g. \citealt[pp. 61]{gillespie_2004}), but other important population ecological assumptions --- such as how density is regulated --- are left unspecified. Beyond being conceptually unsatisfactory \citep{doebeli_2017}, we might also doubt whether this density-independent-$s$, constant-$N$ setup is consistent with the ecology of many populations. Much of the seminal literature on density-dependent selection revolves around the connection between selection in crowded populations and the ability to keep growing at higher densities \citep{macarthur_1967,roughgarden_1979}. The classic example is the logistic model, where the type with the greatest equilibrium population density ``$K$'' excludes the others (Fig.~\ref{fig:Ksel}a). Similarly, the ``$R^*$ rule'', a central tenet of resource competition theory, states that when growth is limited by a single homogeneous consumable resource, the type able to deplete the resource to the lowest equilibrium density $R^*$ (which will often entail a different equilibrium population density) excludes the others \citep{grover_1997}. In these examples, both $N$ and $s$ change during, and as a result of, adaptive sweeps. The Lotka-Volterra competition model also couples selection in crowded populations to density except in special cases \citep{smouse_1976,mallet_2012}. It would therefore seem that the ubiquitous density-independent-$s$, constant-$N$ description of selection is incompatible with a huge class of population ecological processes driving selection (Fig.~\ref{fig:Ksel}b).

\begin{figure}
\centering
\includegraphics[scale=0.8]{Kplot.pdf}
\caption{\label{fig:Ksel} Selection in crowded environments shown as a phase diagram for the densities of two types $n_1$ and $n_2$. (a) The logistic model $\frac{dn_1}{dt}=r_1(1-\frac{n_1+n_2}{K_1})n_1$ and $\frac{dn_2}{dt}=r_2(1-\frac{n_1+n_2}{K_2})n_1$ with $r_1=r_2$ and $K_1>K_2$. (b) The constant-$N$, relative fitness description of selection.}
\end{figure}

In light of this, and the related difficulties posed by frequency-dependent selection, relative fitness values that are independent of type frequencies and population density are typically justified as an approximation that holds when selection is weak (\citealt{kimura1969natural}; \citealt[pp. 277]{ewens_2004}; \citealt[Chap. 4]{charlesworth_1994}) and density is changing much slower than type frequencies \citep[Fig. 9.5]{otto_2011}. Under these conditions, the selection coefficient may depend on density and type frequencies, but since these change slowly the selection coefficient may remain approximately constant over several generations. In the case of density-dependence, this time interval of constancy could include entire selective sweeps. On the other hand, it is known in both simple models like the logistic \citep{kimura1969natural}, and in complex age-structured models \citep[Chap. 4]{charlesworth_1994}, that frequency trajectories under strong density- or frequency-dependent selection can differ substantially from the constant-$s$ approximation.

Yet knowing that the constant-$s$ approximation can fail \textit{in general} when selection is strong gives only a rough picture of the domain and manner in which the constant-$s$ approximation breaks down. For instance, wild \textit{Drosophila} is an important model system in genetics, and exhibits strong seasonally-alternating selection in unison with large ``boom-bust'' density cycles \citep{messer_2016,bergland_14}. Are we compelled to switch to a more ecologically-detailed evolutionary model based on Malthusian parameters or birth/death rates in this system? And if we make this switch, how much ecological detail do we need? Perhaps the strongest justification of constant relative fitness models is that they are a powerful idealization. Thus, before abandoning them it is important to have a better understanding when and how we should do so when selection is not weak, and demography is not stable.

The rough picture given by the weak-selection approximation is missing at least one key element: virtually all models of density-dependent selection do not have a ``reproductive excess'' of juveniles that are more fragile than their adult counterparts \citep{turner1968population,kimura1969natural,nei1971fertility}. Reproductive excesses allow selection to be concentrated at a juvenile ``bottleneck'', which means that the density of adults can remain constant even if strong selection is occurring on juveniles --- rather different from the classic $K$-selection scenario. Although similar ideas featured prominently in   early debates on the regulation of population density (coincidentally grounded in experiments on flies; \citealt{nicholson_1954}), much of the reproductive excess literature revolves around Haldane's ``cost of selection'' \citep{haldane_1957}, and is also poorly integrated with population ecology. For instance, \cite{kimura1969natural} simply took constant $N$ as a requirement and then derived variants of the logistic model consistent with this. \cite{nei1971fertility} proposed a model with an explicit representation of reproductive excess, but used an unusual model of competition based on pair-wise interactions which was only defined for at most two different types. As a result, the role of reproductive excesses in justifying the use of constant relative finesses is largely verbal.

Here we study the population ecology of relative fitness using a novel model of density-dependent population growth based on territorial contests. We restrict our attention to asexual haploids with little or no age structure, since it is then easier to evaluate how the success or failure of the constant-$s$ description is tied to the underlying population ecological assumptions. Our starting point is the classic lottery model which was developed by ecologists to study competition driven by territorial contests in reef fishes and plants \citep{sale_77,chesson_1981}. The classic lottery incorporates a reproductive excess, and fitness involves a product of fertility and juvenile viability akin to standard population genetic models of selection (e.g. \citealt[pp. 185]{crow_1970}). The classic lottery also assumes constant $N$, and generations can overlap. Our first task is to generalize the lottery model to create a variable-density lottery (sections ``Model'' and ``Analytical approximation of the variable-density lottery'').

Equipped with our new variable-density lottery model, we outline its selection properties (section ``Density-dependent selection in the variable-density lottery''). Next, we discuss
selection on the ability to contest territories, which behaves like a pure constant-$N$, relative fitness trait, and discuss how this fits with the classical notion of $K$-selection  (section ''The response of density to selection; $c$-selection versus $K$-selection''). We then consider selection on density-regulating traits (section ``Density-regulating traits and the threat of strong selection''), and conclude by contrasting the classical density-dependent selection literature with our results (``Discussion'').

@@

%Previous models of selection-independent density have either relied on unusual models of competition \citep{kimura1969natural,nei1971fertility}, or made restrictive Lotka-Volterra parameter choices (Appendix C; \citealt{smouse_1976,mallet_2012}).

Summarizing the three traits in the variable-density lottery model: (i) $c$-selection is density-dependent, but $c$ does not regulate density; (ii) $d$ regulates density, but $d$-selection is density-independent; (iii) $b$ regulates density and $b$-selection is density-dependent. Yet despite these differences, pure $b$, $c$ and $d$ sweeps starting and ending at equilibrium all obey the density-independent relative fitness description of selection almost exactly. This behavior is quite different from that found in the classical density-dependent selection literature \citep{roughgarden_1979,christiansen_2004}.

To briefly review: based on a diploid, bi-allelic variant of the logistic model, the $r$/$K$ scheme proposed a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded) \citep{macarthur_1962}, with the latter taken to mean selection for greater equilibrium density \citep{gill_1974}. A more general Lotka-Volterra model introduces the inter-type $\alpha_{ij}$ competition coefficients, with selection on these termed ``$\alpha$-selection'' \citep{gill_1974,joshi_2001}. Setting aside $r$ which confers no selective advantage at equilibrium, we are left with $K$ and $\alpha$, which both behave like $\delta$ in Eq.~\eqref{eq:simplebirthdeath} in that they are density-dependent and cause density to change over a sweep (although $N$ only dips transiently during an $\alpha$-sweep). Thus, strong selection is sufficient for relative fitness to break down in the classical view of density-dependent selection.

Our variable-density lottery model shows that it is not simply a lack of ecological realism that underlies the contrast between relative fitness and the classical view of density-dependent selection. Rather, in many population growth models, only one life-history stage is represented, and the competitive effects resulting from crowding appear as a reduction in absolute fitness that only depends on the type densities at this life-history stage (e.g. the $n_i^2$ and $n_in_j$ terms in the Lotka Volterra equation). As noted in the introduction, this precludes selection concentrated at a fragile juvenile stage as a result of a reproductive excess \citep{chesson_1983,turner1968population,kimura1969natural,nei1971fertility}.

Reproductive excesses appear in the variable-density lottery model when the number of propagules is greater than the number of available territories. Then only $\approx 1/L$ of the juveniles contesting available territories survive to adulthood. Unlike the role of adult density $n_i$ in single-life-stage models, it is the propagule densities $l_i$ that represent the crowding that drives competition (a ``critical age-group''; \citealt[pp. 54]{charlesworth_1994}). In general, reproductive excesses will tend to produce strictly-relative lottery-type contests in which fitter types grow at the expense of others by preferentially filling the available adult ``slots''. The number of slots can remain fixed or change independently of selection at the juvenile stage. By ignoring reproductive excesses, single life-stage models are biased to have total population density be sensitive to ongoing selection. In this respect, the viability selection heuristics that are common in population genetics \cite[pp. 61]{gillespie_2004} actually capture an important ecological process without jumping to the full complexity of age-structured models.

We now turn to the breakdown of the density-independent relative fitness assumption which occurs when strong selection changes population density and is also density-dependent (Fig.~\ref{fig:strengthofselection}). In the variable-density lottery, this occurs if and only if types differ in more than one trait (Fig.~\ref{fig:multiple}). The $c$ and $d$ traits represent the two distinct directions in which density and selection interact: selection may depend on density, and density may depend on selection \citep{prout_1980}. The combination is necessary to pose a threat to Eq.~\eqref{eq:canonical}. However, the $b$ trait remarkably demonstrates that the combination is not sufficient, since the density-dependence of $b$-selection disappears over equilibrium-to-equilibrium $b$-sweeps. Thus, the simple differential/difference equations that have become standard in discussions of density-dependent selection \citep{roughgarden_1979,christiansen_2004,mallet_2012,travis_2013} actually represent a confounded form of the interaction between density and selection.

While this is a conceptual reason to be wary of the classical density-dependent selection models, it is not clear what trait variation to expect in nature. Should we expect mutations to generally affect $b$, $c$ and $d$ independently of each other, or pleiotropically such that $\delta$-like selection is prevalent? In the case of well-mixed indirect exploitation competition for consumable resources, the $R^*$ rule  suggests that $\delta$-like selection will be prevalent. However, for many populations consumable resources are not well-mixed. Spatial localization of consumable resources (e.g. due to restricted movement of nutrients through soils) will tend to create a territorial situation similar to the lottery model, where resource competition only occurs locally and both it and interference competition are subsumed into the competitive ability $c$, which does not affect $N$.

The above findings underscore that the most serious threat to relative fitness arises from deviations from demographic equilibrium. This can dramatically alter frequency trajectories for individual sweeps (e.g. Fig. 9.5 in \cite{otto_2011}; Fig. 5 in \cite{mallet_2012}), as well as the long-term outcomes of selection in fluctuating environments \citep{lande_2009}. This would suggest that in the wild \textit{Drosophila} example mentioned in the Introduction, there is indeed no choice but to abandon relative fitness. The variable-density lottery could provide a useful starting point for analyzing evolution in this and other far-from-equilibrium situations. One convenient feature of our variable-density lottery (and lottery models in general) is a formal similarity to the Wright-Fisher model, which should facilitate the analysis of ``density-dependent drift''.

Another issue with the constant-$N$ relative fitness description of selection is that it precludes consideration of longer-term aspects of the interplay between evolution and ecology such as population extinction. A variety of approaches have been developed for dealing with these issues in quantitative genetics \citep{burger1995evolution,engen_2013}, population genetics \citep{bertram2017predicting} and adaptive dynamics \citep{ferriere2013eco,dieckmann2004adaptive}. Although density-dependent selection is  pertinent to these longer-term issues \citep{travis_2013}, our focus here has been the description of the time-dependent process by which selection changes allele frequencies. This is particularly critical for making sense of evolution at the genetic level, for which we now have abundant data.

>>>>

For haploids growing in discrete time steps, the change in the abundance $n_i$ of type $i$ over a time step can be expressed as $\Delta n_i=(W_i - 1)n_i$ where $W_i$ is ``absolute fitness''. The corresponding change in frequency is $\Delta p_i=\left(\frac{W_i}{\overline{W}}-1\right) p_i$, where $\overline{W}=\sum_i W_i p_i$. In continuous time, the Malthusian parameter $r_i$ replaces $W_i$ and we have $\frac{d n_i}{dt}=r_in_i$ and $\frac{d p_i}{dt}=(r_i-\overline{r}) p_i$ \citep[pp. 26]{crow_1970}. In the particular case that two types are present, the selection equation for the first type takes the familiar form
\begin{equation}
\frac{d p_1}{dt}=s p_1(1-p_1), \label{eq:canonical}
\end{equation}
where $s=r_1-r_2$ is the selection coefficient. Note that, in the discrete case, we can replace the $W_i$ with any set of values proportional to the $W_i$ without affecting the  ratio $W_i/\overline{W}$ or $\Delta p_i$. These ``relative fitness'' values tell us how type frequencies change, but give no information about the dynamics of total population density $N=\sum_i n_i$ \citep[pp. 468]{barton_2007}. Similarly, the selection coefficient and frequency dynamics in Eq.~\eqref{eq:canonical} would be unaffected if we were to add an arbitrary constant to $r_1$ and $r_2$ (these would be relative log fitnesses).

In a constant environment, and in the absence of crowding, $r_i$ is a constant ``intrinsic growth rate (for concreteness we will use Eq.~\eqref{eq:canonical} as our point of comparison, but the discrete time case is similar). The selection coefficient $s$ is then simply the difference in intrinsic population growth rates. The situation is more complicated when we allow for population crowding. Since crowded and uncrowded conditions can favor quite different traits \citep{travis_2013}, the Malthusian parameters $r_i$ may respond differently to changes in $N$. The selection coefficient would then be density-dependent, and we would need to also know the dynamics of $N$ to use Eq.~\eqref{eq:canonical}. In general we cannot just independently specify the dynamics of $N$, because those ecological dynamics are coupled with the evolutionary dynamics of type frequency \citep{travis_2013}. Thus, in a general crowded population the simple procedure of assigning constant relative fitness values to different types has to be replaced with an ecological description of absolute growth rates. Note that frequency-dependent selection does not raise such a severe problem, because we only need to model the type frequencies, not also the ecological variable $N$.

In practice, many population genetics models assume that $N$ has reached some fixed equilibrium value, and each type is assigned a constant relative fitness. Selection is typically assumed to operate through viability (e.g. \citealt[pp. 61]{gillespie_2004}), but other important population ecological assumptions --- such as how density is regulated --- are left unspecified. Beyond being conceptually unsatisfactory \citep{doebeli_2017}, we might also doubt whether this density-independent-$s$, constant-$N$ setup is consistent with the ecology of many populations. Much of the seminal literature on density-dependent selection revolves around the connection between selection in crowded populations and the ability to keep growing at higher densities \citep{macarthur_1967,roughgarden_1979}. The classic example is the logistic model, where the type with the greatest equilibrium population density ``$K$'' excludes the others (Fig.~\ref{fig:Ksel}a). Similarly, the ``$R^*$ rule'', a central tenet of resource competition theory, states that when growth is limited by a single homogeneous consumable resource, the type able to deplete the resource to the lowest equilibrium density $R^*$ (which will often entail a different equilibrium population density) excludes the others \citep{grover_1997}. In these examples, both $N$ and $s$ change during, and as a result of, adaptive sweeps. The Lotka-Volterra competition model also couples density to selection in crowded populations except in special cases \citep{smouse_1976,mallet_2012}. It would therefore seem that the ubiquitous density-independent-$s$, constant-$N$ setup is incompatible with a broad class of population ecological processes driving selection (Fig.~\ref{fig:Ksel}b).

\begin{figure}
\centering
\includegraphics[scale=0.8]{Kplot.pdf}
\caption{\label{fig:Ksel} Phase diagram for the densities of two types $n_1$ and $n_2$ undergoing selection. (a) The logistic model $\frac{dn_1}{dt}=r_1(1-\frac{n_1+n_2}{K_1})n_1$ and $\frac{dn_2}{dt}=r_2(1-\frac{n_1+n_2}{K_2})n_1$ with $r_1=r_2$ and $K_1>K_2$. (b) The constant-$N$, relative fitness description of selection.}
\end{figure}

In light of these difficulties, the assignment of a density- and frequency-independent relative fitness value to each type is typically justified as an approximation that holds when selection is weak (\citealt{kimura1969natural}; \citealt[pp. 277]{ewens_2004}; \citealt[Chap. 4]{charlesworth_1994}) and $N$ changes slowly \citep[Fig. 9.5]{otto_2011}. Under these conditions, $s$ is approximately constant in Eq.~\eqref{eq:canonical}, at least for some number of generations. If $s$ depends only on density, not frequency, this approximate constancy can hold over entire selective sweeps.

However, our understanding of the strong-selection, unstable-$N$ regime is generally limited to a few prominent examples such as the classical density-dependent selection literature discussed above, and complex age-structured models \citep[Chap. 4]{charlesworth_1994}. Since constant relative fitness models are such a powerful idealization, it is important to understand the specifics of how and when they fail when selection is not weak, and $N$ is not stable. For instance, in wild \textit{Drosophila}, strong seasonally-alternating selection happens concurrently with large ``boom-bust'' density cycles \citep{messer_2016,bergland_14}. Are we compelled to switch to a more ecologically-detailed model of selection based on Malthusian parameters or birth/death rates in this important model system? And if we make this switch, how much ecological detail do we need?

Most models of density-dependent selection are missing one key element that has important ramifications for the validity of the constant-$s$ approximation: they do not have a ``reproductive excess'' of juveniles that are more fragile than their adult counterparts \citep{turner1968population}. Reproductive excesses allow selection to be concentrated at a juvenile ``bottleneck'', which means that the density of adults can remain constant even if strong selection is occurring on juveniles --- rather different from the classic $K$-selection scenario. Although similar ideas featured prominently in   early debates on the regulation of population density (grounded in experiments on flies; \citealt{nicholson_1954}), much of the reproductive excess literature revolves around Haldane's ``cost of selection'' \citep{haldane_1957}, and is also poorly integrated with population ecology. For instance, \cite{kimura1969natural} simply took constant $N$ as a requirement and then derived variants of the logistic model consistent with this. \cite{nei1971fertility} proposed a model with an explicit representation of reproductive excess, but used an unusual model of competition based on pair-wise interactions which was only defined for at most two different types. As a result, the role of reproductive excesses in justifying the use of constant relative finesses is largely verbal.

Here we study the population ecology of relative fitness using a novel model of density-dependent population growth based on territorial contests. We restrict our attention to asexual haploids, since it is then clearer how the success or failure of the constant-$s$ description is tied to the underlying population ecological assumptions. Our starting point is the classic lottery model which was developed by ecologists to study competition driven by territorial contests in reef fishes and plants \citep{sale_77,chesson_1981}. The classic lottery incorporates a reproductive excess, and fitness involves a product of fertility and juvenile viability akin to standard population genetic models of selection (e.g. \citealt[pp. 185]{crow_1970}). The classic lottery also assumes constant $N$, and generations can overlap. Our first task is to relax the constant-$N$ requirement of the lottery model to create a variable-density lottery (sections ``Model'' and ``Analytical approximation of the variable-density lottery'').

We then outline the selection properties of our new variable-density lottery model (section ``Density-dependent selection in the variable-density lottery''). Next, we discuss
selection on the ability to contest territories, which behaves like a pure constant-$N$, constant-$s$ trait, and discuss how this fits with the classical notion of $K$-selection  (section ''The response of density to selection; $c$-selection versus $K$-selection''). We then consider selection on traits that regulate population density (section ``Density-regulating traits and the threat of strong selection''), and conclude by contrasting the classical density-dependent selection literature with our results (``Discussion'').

@@

Summarizing the three traits in the variable-density lottery model: (i) $c$-selection is density-dependent, but $c$ does not regulate density; (ii) $d$ regulates density, but $d$-selection is density-independent; (iii) $b$ regulates density and $b$-selection is density-dependent. Yet despite their differences, pure $b$, $c$ and $d$ sweeps starting and ending at equilibrium all obey the density-independent relative fitness description of selection almost exactly.

This behavior is quite different from that found in the classical density-dependent selection literature \citep{roughgarden_1979,christiansen_2004}, where strong selection is sufficient for relative fitness to break down. To briefly review: based on a diploid, bi-allelic variant of the logistic model, the $r$/$K$ scheme proposed a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded) \citep{macarthur_1962}, with the latter taken to mean selection for greater equilibrium density \citep{gill_1974}. The more general Lotka-Volterra competition model introduces the inter-type $\alpha_{ij}$ competition coefficients, with selection on these termed ``$\alpha$-selection'' \citep{gill_1974,joshi_2001}. Setting aside $r$ which confers no selective advantage at equilibrium, we are left with $K$ and $\alpha$. These traits both behave like $\delta$ in Eq.~\eqref{eq:simplebirthdeath} in that they are density-dependent and cause density to change over a sweep (Fig.~\ref{fig:strengthofselection}).

In the variable-density lottery, this occurs if and only if types differ in more than one trait (Fig.~\ref{fig:multiple}). The $c$ and $d$ traits represent the two distinct directions in which density and selection interact: selection may depend on density, and density may depend on selection \citep{prout_1980}. The combination of both is necessary to threaten the constant-$s$ approximation. However, remarkably, the $b$ trait demonstrates that the combination is not sufficient, since the density-dependence of $b$-selection effectively disappears over equilibrium-to-equilibrium $b$-sweeps. Thus, the simple differential/difference equations that have become standard in discussions of density-dependent selection \citep{roughgarden_1979,christiansen_2004,mallet_2012,travis_2013} actually confound important aspects of the interaction between density and selection.

While this is a conceptual reason to be wary of the classical density-dependent selection models, it is not clear what trait variation should be expected in nature. Should we expect mutations to generally affect $b$, $c$ and $d$ independently of each other, or pleiotropically such that $\delta$-like selection is prevalent? In the case of well-mixed indirect exploitation competition for consumable resources, the $R^*$ rule  suggests that $\delta$-like selection will be prevalent. However, for many populations consumable resources are not well-mixed. Spatial localization of consumable resources (e.g. due to restricted movement of nutrients through soils) will tend to create a territorial situation similar to the lottery model, where resource competition only occurs locally and both it and interference competition are subsumed into the competitive ability $c$, which does not affect $N$.

Our variable-density lottery model shows that it is not simply a lack of ecological realism on the part of constant relative fitness models that underlies its contrast with the classical view of density-dependent selection. Rather, in many population growth models, only one life-history stage is represented, and the competitive effects resulting from crowding appear as a reduction in absolute fitness that only depends on the type densities at this life-history stage (e.g. the $n_i^2$ and $n_in_j$ terms in the Lotka Volterra equation). As This precludes selection concentrated at a fragile juvenile stage facilitated by a reproductive excess \citep{chesson_1983,turner1968population,kimura1969natural,nei1971fertility}.

Reproductive excesses appear in the variable-density lottery model when the number of propagules is greater than the number of available territories. Then only $\approx 1/L$ of the juveniles contesting unoccupied territories survive to adulthood. Unlike the role of adult density $n_i$ in single-life-stage models, it is the propagule densities $l_i$ that represent the crowding that drives competition (a ``critical age-group''; \citealt[pp. 54]{charlesworth_1994}). In general, reproductive excesses will tend to produce strictly-relative lottery-type contests in which fitter types grow at the expense of others by preferentially filling the available adult ``slots''. The number of slots can remain fixed or change independently of selection at the juvenile stage. By ignoring reproductive excesses, single life-stage models are biased to have total population density be sensitive to ongoing selection. In this respect, the viability selection heuristics that are common in population genetics \cite[pp. 61]{gillespie_2004} actually capture an important ecological process without making the full leap to complex age-structured models.

The above findings underscore that the most serious threat to the constant-$s$ approximation arises due to deviations from demographic equilibrium as a result of changes in the demographic rates of the types already present i.e. as a result of a temporally-variable environment. While transient deviations from demographic equilibrium driven by the appearance of new types can also threaten the constant-$s$ approximation, it requires strong selection that is density-dependent and effects a density-regulating trait (and even then the constant-$s$ approximation may hold). Temporally-variable environments can dramatically alter frequency trajectories for individual sweeps (e.g. Fig. 9.5 in \cite{otto_2011}; Fig. 5 in \cite{mallet_2012}), as well as the long-term outcomes of selection \citep{lande_2009}.

This suggests that in systems like the wild \textit{Drosophila} example mentioned in the Introduction, there is indeed no choice but to abandon relative fitness. The variable-density lottery could provide a useful starting point for analyzing evolution in this and other far-from-equilibrium situations for two reasons: 1) the $b$, $c$, $d$ trait scheme neatly distinguishes between different aspects of the interplay between density and selection; 2)  lottery models in general are mathematically similarity to the Wright-Fisher model, which should facilitate the analysis of genetic when $N$ is unstable.

[Kind of an odd way to end, but it breaks the flow if I put it elsewhere] Another issue with the constant-$N$, constant-$s$ description of selection is that it precludes consideration of longer-term aspects of the interplay between evolution and ecology such as population extinction. A variety of approaches have been developed for dealing with these issues in quantitative genetics \citep{burger1995evolution,engen_2013}, population genetics \citep{bertram2017predicting} and adaptive dynamics \citep{ferriere2013eco,dieckmann2004adaptive}. Although density-dependent selection is  pertinent to these longer-term issues \citep{travis_2013}, our focus here has been the description of the time-dependent process by which selection changes allele frequencies. This is particularly critical for making sense of evolution at the genetic level, for which we now have abundant data.
